\LARGE\textbf{Designing Studies and Evaluating Research Results:\\ Type M and Type S Errors for\\ Pearson Correlation Coefficient }

### 2021-03-16 Documents/Paper_main/Paper_main.Rnw
revision altoe

Psychological science is increasingly committed to scrutinizing its published findings by promoting large-scale replication efforts, where the protocol of a previous study is repeated as closely as possible with a new sample\parencite{camererEvaluatingReplicabilityLaboratory2016, camererEvaluatingReplicabilitySocial2018, ebersoleManyLabsEvaluating2016, kleinInvestigatingVariationReplicability2014, kleinManyLabsInvestigating2018, opensciencecollaborationEstimatingReproducibilityPsychological2015}. Interestingly, many replication studies found smaller effects than originals \parencite{camererEvaluatingReplicabilitySocial2018, opensciencecollaborationEstimatingReproducibilityPsychological2015} and among many possible explanations, one relates to a feature of study design: statistical power. In particular, it is plausible for original studies to have lower statistical power than their replications. In the case of underpowered studies, we are usually aware of the lower probability of detecting an effect if this exists, but the less obvious consequences on effect size estimation are often neglected. When underpowered studies are analyzed using thresholds, such as statistical significance levels, effects passing such thresholds have to exaggerate the true effect size \parencite{buttonPowerFailureWhy2013, gelmanTypeErrorMight2017, ioannidisWhyMostDiscovered2008, ioannidisEmergenceLargeTreatment2013, laneEstimatingEffectSize1978}. Indeed, as will be extensively shown below, in underpowered studies only large effects correspond to values that can reject the null hypothesis and be statistically significant. As a consequence, if the original study was underpowered and found an exaggerated estimate of the effect, the replication effect will likely be smaller.

@@

Why do these errors matter? The concepts of Type M and Type S errors allow enhancing researchers’ awareness of a complex process such as statistical inference. Strictly speaking, Design Analysis used in the design phase of a study provides similar information as the classical power analysis, indeed, to a given level of power there is a corresponding Type M and Type S errors. However, it is a valuable conceptual framework that can help researchers to understand the important role of statistical power both when designing a new study or when evaluating previous results from the literature. In particular, it highlights the unwanted (and often overlooked) consequences on effect estimation when filtering for statistical significance in underpowered studies. In these scenarios, there is not only a lower probability of rejecting the null when it is actually false but, even more importantly, any significant result would most likely lead to a misleading overestimation of the actual effect. The exaggeration of effect sizes, in the right or the wrong direction, has important implications on a theoretical and applied level. On a theoretical level, studies’ designs with high Type M and Type S errors can foster distorted expectations on the effect under study, triggering a vicious cycle for the planning of future studies. This point is relevant also for the design of replication studies, where the possible inflation of the original effect should be taken into account together with the smallest effect size of interest to limit the chances of being underpowered \parencite{buttonPowerFailureWhy2013}. When studies are used to inform policymaking and real-world interventions, implications can go beyond the academic research community and can impact society at large. In these settings, we could assist to a “hype and disappointment cycle” \parencite{gelmanOverconfidenceResearchCertainty2019}, where true effects turn out to be much less impressive than expected. This can produce undesirable consequences on people’s lives, a consideration that invites researchers to assume responsibility in effectively communicating the risks related to effects quantification.

@@

The central and possibly most difficult point in this process is reasoning on what could be the plausible magnitude and direction of the effect of interest. This critical process, which is central also in traditional power analysis, is an opportunity for researchers to aggregate, formalize and incorporate prior information on the phenomenon under investigation \parencite{gelmanPowerCalculationsAssessing2014}. What is plausible can be determined on theoretical grounds, using expert knowledge elicitation techniques and consulting literature reviews and meta-analysis, always taking into account the presence of effect sizes inflation in the published literature \parencite{andersonBestOftForgotten2019}. The plausible effect size approximates the true effect, which is never known but can be thought of as “that which would be observed in a hypothetical infinitely large sample” \parencite[p. 642]{gelmanPowerCalculationsAssessing2014}. For a more exhaustive description of plausible effect size, we refer the interested reader to \textcite{altoeEnhancingStatisticalInference2020, gelmanPowerCalculationsAssessing2014}.

@@

In the following sections, we illustrate how to perform prospective and retrospective design analysis using some examples. We developed two R functions\footnote{An R-package was subsequently developed and now is available on CRAN, PRDA: Conduct a Prospective or Retrospective Design Analysis \url{https://cran.r-project.org/web/packages/PRDA/index.html}.} to perform design analysis for Pearson correlation, which are available at the page \url{https://osf.io/9q5fr/}. The function to perform a prospective design analysis is \texttt{pro\_r()}. It requires as input the plausible effect size (\texttt{rho}), the statistical power (\texttt{power}), the directionality of the test (\texttt{alternative}) which can be set as: “\texttt{two.sided}”, “\texttt{less}” or “\texttt{greater}”. Type I error rate (\texttt{sig\_level}) is set as default at 5\% and can be changed by the user. The \texttt{pro\_r()} function returns the necessary sample size to achieve the desired statistical power, Type M error rate, the Type S error probability, and the critical value(s) above which a statistically significant result can be found. The function to perform retrospective design analysis is \texttt{retro\_r()}. It requires as input the plausible effect size, the sample size used in the study, and the directionality of the test that was performed. Also in this case, Type I error rate is set as default at 5\% and can be changed by the user.  The function \texttt{retro\_r()} returns the Type M error rate, the Type S error probability, and the critical value(s)\footnote{\emph{Critical value} is the name usually employed in hypotheses testing within the Neyman-Pearson framework. In the research practice, this is also known as the \emph{Minimal Statistically Detectable Effect} \parencite{cookAssessingMethodsSpecify2014, phillipsStatisticalSignificanceSediment2001}}. For further details regarding the R functions refer to Appendix~\ref{sec:appendix_b}.

@@

Imagine that we were part of the research team in the previous case study exploring the relationship between activity in the Anterior Cerebral Cortex and perceived distress. When drafting the research protocol, we face the inevitable discussion on how many participants we are going to recruit. This choice depends on available resources, type of study design, constraints of various nature and, importantly, the plausible magnitude and direction of the phenomenon that we are going to study or the smallest effect size of interest. As previously mentioned, deciding on a plausible effect size is a fundamental step which requires great effort and should not be done by trying different values only to obtain a more desirable sample size. Instead, proposing a plausible effect size is where the expert knowledge of the researcher can be formalized and can greatly contribute to the informativeness of the study that is being planned. For the sake of these examples, we adopt the previous consideration and we suppose that common agreement is reached on a plausible correlation coefficient to be around $\rho=.25$. Finally, we would like to leave open the possibility to explore whether the relationship goes in the opposite direction to the one hypothesized, so we decide to perform a two-tailed test.

@@

The critical element is the application of the statistical significance filter without taking into account statistical power. Design Analysis per se does not solve this issue but, instead, it allows us to recognize its problematic consequences. In the same way as statistical power is a characteristic of a study design, so are Type M and Type S errors, however, the two are qualitatively different in terms of the kind of reasoning they favour. Statistical power is defined in terms of probability of rejecting the Null hypothesis and, even though this is based on an effect size of interest, the relationship “low power - high possibility of exaggeration” may not be straightforward for everyone. Instead, Type M and Type S errors directly quantify the possible exaggeration. Furthermore, their consideration protects against another possible pitfall. It is quite widespread the knowledge that statistically significant results can be not-meaningful when they are associated with small effect sizes. However, if a statistically significant result is found paired with a large effect size, the results may be considered robust. The problem is that this is not true in all cases. Here, the missing piece of information is  statistical power. If power is considered, researchers would realize that a large effect was found in a context where there was a low probability to find it. But this interpretation is not explicitly stating an important aspect: in these conditions, the only way to find a statistically significant result is by overestimating the true effect. On the contrary, this consequence becomes immediately clear once Type M and Type S errors are considered retrospectively. Similarly, considering Type M and Type S prospectively favours reasoning in terms of effect size rather than the probability of rejecting the null hypothesis when setting the sample size in a design analysis.

@@

>>>>

Psychological science is increasingly committed to scrutinizing its published findings by promoting large-scale replication efforts, where the protocol of a previous study is repeated as closely as possible with a new sample \parencite{camererEvaluatingReplicabilityLaboratory2016, camererEvaluatingReplicabilitySocial2018, ebersoleManyLabsEvaluating2016, kleinInvestigatingVariationReplicability2014, kleinManyLabsInvestigating2018, opensciencecollaborationEstimatingReproducibilityPsychological2015}. Interestingly, many replication studies found smaller effects than originals \parencite{camererEvaluatingReplicabilitySocial2018, opensciencecollaborationEstimatingReproducibilityPsychological2015} and among many possible explanations, one relates to a feature of study design: statistical power. In particular, it is plausible for original studies to have lower statistical power than their replications. In the case of underpowered studies, we are usually aware of the lower probability of detecting an effect if this exists, but the less obvious consequences on effect size estimation are often neglected. When underpowered studies are analyzed using thresholds, such as statistical significance levels, effects passing such thresholds have to exaggerate the true effect size \parencite{buttonPowerFailureWhy2013, gelmanTypeErrorMight2017, ioannidisWhyMostDiscovered2008, ioannidisEmergenceLargeTreatment2013, laneEstimatingEffectSize1978}. Indeed, as will be extensively shown below, in underpowered studies only large effects correspond to values that can reject the null hypothesis and be statistically significant. As a consequence, if the original study was underpowered and found an exaggerated estimate of the effect, the replication effect will likely be smaller.

@@

Why do these errors matter? The concepts of Type M and Type S errors allow enhancing researchers’ awareness of a complex process such as statistical inference. Strictly speaking, Design Analysis used in the design phase of a study provides similar information as the classical power analysis, indeed, to a given level of power there is a corresponding Type M and Type S errors. However, it is a valuable conceptual framework that can help researchers to understand the important role of statistical power both when designing a new study or when evaluating previous results from the literature. In particular, it highlights the unwanted (and often overlooked) consequences on effect estimation when filtering for statistical significance in underpowered studies. In these scenarios, there is not only a lower probability of rejecting the null when it is actually false but, even more importantly, any significant result would most likely lead to a misleading overestimation of the actual effect. The exaggeration of effect sizes, in the right or the wrong direction, has important implications on a theoretical and applied level. On a theoretical level, studies’ designs with high Type M and Type S errors can foster distorted expectations on the effect under study, triggering a vicious cycle for the planning of future studies. This point is relevant also for the design of replication studies, which could turn out to be underpowered if they do not take into account possible inflations of the original effect \parencite{buttonPowerFailureWhy2013}. When studies are used to inform policymaking and real-world interventions, implications can go beyond the academic research community and can impact society at large. In these settings, we could assist to a “hype and disappointment cycle” \parencite{gelmanOverconfidenceResearchCertainty2019}, where true effects turn out to be much less impressive than expected. This can produce undesirable consequences on people’s lives, a consideration that invites researchers to assume responsibility in effectively communicating the risks related to effects quantification.

@@

The central and possibly most difficult point in this process is reasoning on what could be the plausible magnitude and direction of the effect of interest. This critical process, which is central also in traditional power analysis, is an opportunity for researchers to aggregate, formalize and incorporate prior information on the phenomenon under investigation \parencite{gelmanPowerCalculationsAssessing2014}. What is plausible can be determined on theoretical grounds, using expert knowledge elicitation techniques \parencite[see for example][]{ohaganExpertKnowledgeElicitation2019} and consulting literature reviews and meta-analysis, always taking into account the presence of effect sizes inflation in the published literature \parencite{andersonBestOftForgotten2019}. Given these premises, it is important to stress that a plausible effect size should not be determined by considering the results of a single study, given the high-level of uncertainty associated with his effect size estimate. The idea is that the plausible effect size should approximate the true effect, which - although never known - can be thought of as “that which would be observed in a hypothetical infinitely large sample” \parencite[p. 642]{gelmanPowerCalculationsAssessing2014}. For a more exhaustive description of plausible effect size, we refer the interested reader to \textcite{altoeEnhancingStatisticalInference2020, gelmanPowerCalculationsAssessing2014}.

Before we proceed, it is worth noting that there are other recent valuable tools that start from different premises for designing and evaluating studies. Among others, we refer the interested reader to methods which start from the definition of a small effect size of interest \parencite[SESOI; for a tutorial, see][]{lakensEquivalenceTestingPsychological2018}.

@@

In the following sections, we illustrate how to perform prospective and retrospective design analysis using some examples. We developed two R functions\footnote{An R-package was subsequently developed and now is available on CRAN, PRDA: Conduct a Prospective or Retrospective Design Analysis \url{https://cran.r-project.org/web/packages/PRDA/index.html}. PRDA contains  other features on Design Analysis,  that are beyond the aim of the present paper.} to perform design analysis for Pearson correlation, which are available at the page \url{https://osf.io/9q5fr/}. The function to perform a prospective design analysis is \texttt{pro\_r()}. It requires as input the plausible effect size (\texttt{rho}), the statistical power (\texttt{power}), the directionality of the test (\texttt{alternative}) which can be set as: “\texttt{two.sided}”, “\texttt{less}” or “\texttt{greater}”. Type I error rate (\texttt{sig\_level}) is set as default at 5\% and can be changed by the user. The \texttt{pro\_r()} function returns the necessary sample size to achieve the desired statistical power, Type M error rate, the Type S error probability, and the critical value(s) above which a statistically significant result can be found. The function to perform retrospective design analysis is \texttt{retro\_r()}. It requires as input the plausible effect size, the sample size used in the study, and the directionality of the test that was performed. Also in this case, Type I error rate is set as default at 5\% and can be changed by the user.  The function \texttt{retro\_r()} returns the Type M error rate, the Type S error probability, and the critical value(s)\footnote{\emph{Critical value} is the name usually employed in hypotheses testing within the Neyman-Pearson framework. In the research practice, this is also known as the \emph{Minimal Statistically Detectable Effect} \parencite{cookAssessingMethodsSpecify2014, phillipsStatisticalSignificanceSediment2001}}. For further details regarding the R functions refer to Appendix~\ref{sec:appendix_b}.

@@

Imagine that we were part of the research team in the previous case study exploring the relationship between activity in the Anterior Cerebral Cortex and perceived distress. When drafting the research protocol, we face the inevitable discussion on how many participants we are going to recruit. This choice depends on available resources, type of study design, constraints of various nature and, importantly, the plausible magnitude and direction of the phenomenon that we are going to study. As previously mentioned, deciding on a plausible effect size is a fundamental step which requires great effort and should not be done by trying different values only to obtain a more desirable sample size. Instead, proposing a plausible effect size is where the expert knowledge of the researcher can be formalized and can greatly contribute to the informativeness of the study that is being planned. For the sake of these examples, we adopt the previous consideration and we suppose that common agreement is reached on a plausible correlation coefficient to be around $\rho=.25$. Finally, we would like to leave open the possibility to explore whether the relationship goes in the opposite direction to the one hypothesized, so we decide to perform a two-tailed test.

@@

The critical element is the application of the statistical significance filter without taking into account statistical power. Design Analysis per se does not solve this issue but, instead, it allows us to recognize its problematic consequences. In the same way as statistical power is a characteristic of a study design, so are Type M and Type S errors, however, the two are qualitatively different in terms of the kind of reasoning they favour. Statistical power is defined in terms of probability of rejecting the Null hypothesis and, even though this is based on an effect size of interest, the relationship “low power - high possibility of exaggeration” may not be straightforward for everyone. Instead, Type M and Type S errors directly quantify the possible exaggeration. Furthermore, their consideration protects against another possible pitfall. When in a study a statistically significant result is found and the associated effect size estimate is large, the finding could be interpreted as robust and impressive. However, this interpretation is not always appropriate. Here, the missing piece of information is  statistical power. If power is considered, researchers would realize that a large effect was found in a context where there was a low probability to find it. But this interpretation is not explicitly stating an important aspect: in these conditions, the only way to find a statistically significant result is by overestimating the true effect. On the contrary, this consequence becomes immediately clear once Type M and Type S errors are considered retrospectively. Similarly, considering Type M and Type S prospectively favours reasoning in terms of effect size rather than the probability of rejecting the null hypothesis when setting the sample size in a design analysis.

@@

It is important to remark that even if Design Analysis is based on the definition of a plausible effect size, a best practice should be to conduct multiple Design Analyses by considering different scenarios which  include different plausible effect sizes and levels of power to maximize the informativeness of both a prospective and a retrospective analysis.
